---
title: "Tidyverse - Exercises"
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
```

```{r xaringan-themer, include=FALSE, warning=FALSE}
library(tidyverse)
library(palmerpenguins)
data(package = "palmerpenguins")
library(knitr)
library(kableExtra)
```

<br>

- Install the Tidyverse or update it. Then load it.

- Check out magrittr’s help file on the forward pipe operator (`%>%`).

- Let’s compare the Base R approach with the Tidyverse approach:

  a. First, install the package `palmerpenguins` and load the built-in penguins dataset into RStudio. You can do this by typing `data(package = "palmerpenguins")`. How many rows and variables does the `penguins` dataset have? And how many factors?

  b. Now select the following variables: `species`, `island`, `flipper_length_mm`, `sex`, `year`, once with the Base R approach, and once with the Tidyverse approach.

  c. Remove all NAs. Again, once with the Base R and once with the Tidyverse approach.
    
- What’s the median flipper length of the different species? Try to obtain the same results with both approaches.

- What’s the minimum body mass on the different islands? Again, try to find the same results with both approaches.