---
title: "A Brief Intro to Plotting with ggplot2"
subtitle: "code-along"
date: "14.09.2021"
---

```{r include=FALSE, warning =FALSE}
library(tidyverse)
library(palmerpenguins)
library(viridis)
```

# How does ggplot work ?

```{r,  warning =FALSE}
ggplot(penguins) +      
  aes(x=bill_length_mm,  
      y=bill_depth_mm) + 
  geom_point()
```

# Example

### Plot distribution of bill length per islands:

```{r, warning =FALSE}
ggplot(penguins)

```

### A boxplot may hide information which can be made accessible with jitter

```{r, warning =FALSE}

```

### Another possibility is a violin plot (works best with lots of data)

```{r, warning =FALSE}

```

### Combining violin with boxplot is a powerful option

```{r, warning =FALSE}

```

### Let's add title, labels, and a theme

```{r, warning =FALSE}

```

### Let's distinguish between species

```{r, warning =FALSE}

```

### And change colors to the viridis scale (optimized for colorblind, and grayscale printing)

```{r, warning =FALSE}

```

### Rename the legend

```{r, warning =FALSE}

```

### Use the power of faceting

```{r, warning =FALSE}

```

### And save the plot as .png with ggsave().

```{r, warning =FALSE}

```
