---
title: "Plotting with ggplot2"
subtitle: "Final Exercise Solution"
date: "15.09.2021"
output: html_notebook
---

```{r}
library(tidyverse)
library(palmerpenguins)
library(viridis)
```

# 1. Plot the mean body weight of female penguins of different species over time

- Add tilte, labels, and rename the legend, 
- Use the viridis color scale
- Save the plot as "mean_weight_by_species_over_time.png"

#### Here goes your code
```{r}
penguins %>% 
```

# Hints:
```{r}
hints_1 <- read_rds("hints_1.dta")
```

Simply run the lines below with ctrl+enter to get hints
```{r hint_1}
print(hints_1[1])

print(hints_1[2])

print(hints_1[3])

print(hints_1[4])

print(hints_1[5])

print(hints_1[6])

print(hints_1[7])
```


# 2. For the motivated among you let's extend the exercise

- Distinguish between islands
- Add errorbars of +- one standard deviation
- Clean the x-axis i.e. make sure only ticks for the relevant years are shown

# Here goes your code

### Reuse your previous solution
```{r}
penguins %>% 
```
# Hints:
```{r}
hints_2 <- read_rds("hints_2.dta")
```

Simply run the lines below with ctrl+enter to get hints
```{r hint_2}
print(hints_2[1])

print(hints_2[2])

print(hints_2[3])

print(hints_2[4])

print(hints_2[5])

print(hints_2[6])
```