---
title: "Plotting with ggplot2"
subtitle: "Final Exercise Solution"
date: "15.09.2021"
output: html_notebook
---

```{r}
library(tidyverse)
library(palmerpenguins)
library(viridis)
```

# 1. Plot the mean body weight of female penguins of different species over time

- Add tilte, labels, and rename the legend, 
- Use the viridis color scale
- Save the plot as "mean_weight_by_species_over_time.png"

#### Here goes your code
```{r}
penguins %>% 
  group_by(year, species) %>% 
  mutate(mean_weight = mean(body_mass_g, na.rm=T)) %>% 
  filter(sex=="female") %>% 
  ggplot() +
    aes(x=year, y=mean_weight, color=species) +
    geom_line() +
    geom_point(size=2) +
    labs(title="Mean weight of different penguin species from 2007 to 2008",
         x="Year",
         y="Mean body weight (g)",
         color="Species")+
    scale_color_viridis(discrete=T)
ggsave("mean_weight_by_species_over_time.png")
```

# Hints:
```{r}
hints_1 <- list("Use group_by()", "Gropu by year and species", "Use mutate to create a variable for the mean", "Filter for sex", "Use the color aesthetics","Remeber to handle NA's in mean()", "Remember ggsave()")
write_rds(hints_1, file = "hints_1.dta")
hints_1 <- read_rds("hints_1.dta")
```

Simply run the lines below with ctrl+enter to get hints
```{r hint_1}
print(hints_1[1])

print(hints_1[2])

print(hints_1[3])

print(hints_1[4])

print(hints_1[5])

print(hints_1[6])

print(hints_1[7])
```


# 2. For the motivated among you let's extend the exercise

- Distinguish between islands
- Add errorbars of +- one standard deviation
- Clean the x-axis i.e. make sure only ticks for the relevant years are shown

```{r}
penguins %>% 
  group_by(year, species, island) %>% 
  mutate(mean_weight = mean(body_mass_g, na.rm=T), sd_weight = sd(body_mass_g, na.rm = T)) %>% 
  ggplot() +
    aes(x=year, y=mean_weight, color=species) +
    geom_line() +
    geom_point(size=2) +
    geom_errorbar(aes(ymax=mean_weight+sd_weight, ymin=mean_weight-sd_weight), width=0.2) +
    labs(title="Mean weight of different penguin species from 2007 to 2008",
         x="Year",
         y="Mean body weight (g)",
         color="Species")+
    scale_color_viridis(discrete=T) +
    scale_x_continuous(breaks = c(2007, 2008, 2009)) +
    facet_grid(rows = vars(island))
```
# Hints:
```{r}
hints_2 <-list("Use Facet grid for the islands", "Also group by islands!", "You get the standard deviation by using sd()", "geom_errorbar, has a parameter for width", "The x can be influenced through scale_x_*", "Ticks are called breaks in R")
write_rds(hints_2, file = "hints_2.dta")
hints_2 <- read_rds("hints_2.dta")
```

Simply run the lines below with ctrl+enter to get hints
```{r hint_2}
print(hints_2[1])

print(hints_2[2])

print(hints_2[3])

print(hints_2[4])

print(hints_2[5])

print(hints_2[6])
```