---
title: "Welcome to the R Bootcamp 2021!"
output:
  xaringan::moon_reader:
    css: xaringan-themer.css
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(fig.retina = 3, message = FALSE) #rendering sharp plots / not showing messages
```

```{r xaringan-themer, include=FALSE, warning=FALSE}
library(xaringanthemer)
style_mono_light(base_color = "#081c15")
#class: inverse, center, middle
library(tidyverse)
```

## Overall schedule

- Admin

- Meet the instructors

- Day 1: The basics of R (Studio)

- Day 2: Data wrangling and visualizations

- Social event with GessWho!

---


## First things first

- Please (properly) wear masks at all time

- The lessons are fairly packed but don't worry: Everything is available [online](https://github.com/bonschorno/rbootcamp_2021)

- Feel free to ask questions at any time

- We have very different levels when it comes to coding. Please keep this in mind and help first rather than getting impatient

**The guiding teaching principle of this course:**

> Tell me and I forget <br> teach me and I remember, <br> involve me and I learn.

<br>

- Shoutout to Charles Lanfear (@clanfear) for providing the slides

- Don't hesitate to share the slides and exercises as well

---

class: inverse, center, middle

## Meet the instructors

---

## Today's schedule

<br>

|Time       |Content                           |
|-----------|----------------------------------|
|10:00      |Meet & Greet Breakfast            |
|10:45-11:30|General introduction to R (Studio)|
|11:45-12:30|Objects, functions, scripts       |
|12:30-14:00|Lunch Break                       |
|14:00-14:45|Data formats, packages, indexing  |
|15:00-15:45|Basic plots and final exercise    |

---

## Any experience with R?

```{r, include=FALSE}
n <- c(12, 5, 13, 2)
names(n) <- c("No", "Yes, on my own", "Yes, during undergrad", "Yes, for work")

plot_experience <- bind_rows(n) %>% 
  pivot_longer(everything(), names_to = "experience", values_to = "value") %>%
  mutate(freq_rel = value/sum(value)) %>% 
  ggplot(aes(x = fct_reorder(experience, -freq_rel), y = freq_rel)) +
  geom_col() +
  scale_y_continuous(labels = scales::label_percent(accuracy = 1)) +
  theme_xaringan() +
  theme(axis.title = element_blank(),
        panel.grid = element_blank(),
        axis.ticks.x = element_blank())
```

```{r, echo=FALSE, fig.width=10}
plot_experience
```


---

## Any experience with other languages?

```{r, echo=FALSE}
languages <- c(3, 10, 7,8, 3)
names(languages) <- c("Stata", "Python", "Matlab", "SPSS", "Multiple")

plot_languages <- bind_rows(languages) %>% 
  pivot_longer(everything(), names_to = "experience", values_to = "value") %>%
  mutate(freq_rel = value/sum(value)) %>% 
  ggplot(aes(x = fct_reorder(experience, -value), y = value)) +
  geom_col() +
  scale_y_continuous(breaks = seq(0,10,5), labels = seq(0,10,5)) +
  theme_xaringan() +
  theme(axis.title = element_blank(),
        panel.grid = element_blank(),
        axis.ticks.x = element_blank())
```

```{r, echo=FALSE, fig.width=10}
plot_languages
```

---

## Equal parts STP/MACIS students

```{r, echo=FALSE, fig.width=10}
# library
library(treemapify)
 
# Create data
group <- c("STP","MACIS")
value <- c(13,15)
data <- data.frame(group,value)
 
# treemap
ggplot(data, 
aes(area = value, fill = group,label = group)) +
   geom_treemap() +
   geom_treemap_text(
     colour = "white",
     place = "centre",
     size = 15) +
  theme_xaringan() +
  labs(fill = "") +
  scale_fill_brewer(palette = "Reds") +
  theme(legend.position = "none")
```

---

# R or Python?

- The primary objective of R is/was data analysis and statistics. The primary objective of Python is/was deployment and production. 

**BUT:**

- What language do your colleagues use?

- Who is working with the data?

- What tasks are you performing?

---
# Why R?

R is a programming language built for statistical computing.

If one already knows Stata or similar software, why use R?

--

* R is *free*, so you don't need a terminal server or license.

--

* R has a *very* large community for support and packages.

--

* R can handle virtually any data format.

--

* R makes replication *easy*.

--

* R is a *language* so it can do *everything*.<sup>1</sup>

.footnote[[1] Including generate these slides (using RMarkdown)!]

---

# R Studio

R Studio is a "front-end" or integrated development environment (IDE) for R that can make your life *easier*.

--

RStudio can:
--

* Organize your code, output, and plots.

--

* Auto-complete code and highlight syntax.

--

* Help view data and objects.

--

* [Enable easy integration of R code into documents.](https://rmarkdown.rstudio.com/)

---
# Getting Started

Open up RStudio now and choose *File > New File > R Script*.

Then, let's get oriented with the interface:

* *Top Left*: Code **editor** pane, data viewer (browse with tabs)

* *Bottom Left*: **Console** for running code (`>` prompt)

* *Top Right*: List of objects in **environment**, code **history** tab.

* *Bottom Right*: Tabs for browsing files, viewing plots, managing packages, and viewing help files.

You can change the layout in *Preferences > Pane Layout*. Also, there are a lot of very useful shortcuts!

---

# Editing and Running Code

There are several ways to run R code in RStudio:
--

* Highlight lines in the **editor** window and click *Run* at the top or hit `Ctrl+Enter` or `⌘+Enter` to run them all.

--

* With your **caret** on a line you want to run, hit `Ctrl+Enter` or `⌘+Enter`. Note your caret moves to the next line, so you can run code sequentially with repeated presses. 

--

* Type individual lines in the **console** and press `Enter`.

--

The console will show the lines you ran followed by any printed output.

---

# Incomplete Code

If you mess up (e.g. leave off a parenthesis), R might show a `+` sign prompting you to finish the command:

```{r Coding 1, eval=FALSE}
> (11-2
+
```

Finish the command or hit `Esc` to get out of this.

---

# R as a Calculator

In the **console**, type `123 + 456 + 789` and hit `Enter`.
--

```{r Calc 1}
123 + 456 + 789
```

--

The `[1]` in the output indicates the numeric **index** of the first element on that line.

--

Now in your blank R document in the **editor**, try typing the line `sqrt(400)` and either
clicking *Run* or hitting `Ctrl+Enter` or `⌘+Enter`.

--

```{r Calc 2}
sqrt(400)
```

---

# Functions and Help

`sqrt()` is an example of a **function** in R.

If we didn't have a good guess as to what `sqrt()` will do, we can type `?sqrt` in the console
and look at the **Help** panel on the right.

```{r Help, eval=FALSE}
?sqrt
```

**Arguments** are the *inputs* to a function. In this case, the only argument to `sqrt()`
is `x` which can be a number or a vector of numbers.

Help files provide documentation on how to use functions and what functions produce.

---

# R Projects

Usually, you have several scripts, numerous datasets and other files. Where are they stored?

```{r, eval=FALSE}
getwd()
```

You can specify the working directory at the beginning of each session...

```{r, eval=FALSE}
setwd()
```

... or use R Projects (a product by R Studio) instead. How does it work?

1. Creates a project file (with an .Rproj extension) within the project directory.
2. Creates a hidden directory (named .Rproj.user) where project-specific temporary files are stored. 
3. Loads the project into RStudio and display its name in the Projects toolbar 

This is super useful when collaborating on different operating systems!
