---
title: "Objects, dataframes, formats"
output:
  xaringan::moon_reader:
    css: xaringan-themer.css
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      always_allow_html: true

---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(fig.retina = 3, message = FALSE) #rendering sharp plots / not showing messages
```

```{r xaringan-themer, include=FALSE, warning=FALSE}
library(xaringanthemer)
style_mono_light(base_color = "#081c15")
#class: inverse, center, middle
library(tidyverse)
```

# Creating Objects

R stores *everything* as an **object**, including data, functions, models, and output.

--

Creating an object can be done using the **assignment operator**: `<-` 
--

```{r Objects 1}
new.object <- 144
```

--

**Operators** like `<-` are functions that look like symbols but typically sit between their arguments 
(e.g. numbers or objects) instead of having them inside `()` like in `sqrt(x)`. 

Technically, you can also use `=` as an assignment operator (like in Python). However, `=` only works within the same scope.

---

# Example

```{r}
mean(x = 1:5)
```

```{r, error=TRUE}
x
```

```{r}
mean(x <- 1:5)
```

```{r}
x
```

So, stick to `<-` when working in R. 
---

# Calling Objects

You can display or "call" an object simply by using its name.

```{r Objects 2}
new.object
```

--

Object names can contain `_` and `.` in them, but cannot *begin* with numbers. Try
to be consistent in naming objects. RStudio auto-complete means *long names are better 
than vague ones*! 

Put differently:

<blockquote class="twitter-tweet" data-width="550" data-lang="en" data-dnt="true" data-theme="light"><p lang="en" dir="ltr">I promise you that the number of characters you save by randomly removing letters from variable names is far fewer than the number of characters you’ll type in vain mis-remembering those abbreviations</p>&mdash; Emily Riederer (@EmilyRiederer) <a href="https://twitter.com/EmilyRiederer/status/1432800401460453381?ref_src=twsrc%5Etfw">August 31, 2021</a></blockquote>

---

# Using Objects

An object's **name** represents the information stored in that **object**, so you can treat the object's name
as if it were the values stored inside.
--

```{r Objects 3}
new.object + 10
new.object + new.object
sqrt(new.object)
```

---

# Creating Vectors

A **vector** is a series of **elements**, such as numbers. 

--

You can create a vector and store it as an object in the same way. To do this, use the
function `c()` which stands for "combine" or "concatenate".
--

```{r Vectors 1}
new.object <- c(4, 9, 16, 25, 36)
new.object
```

--

If you name an object the same name as an existing object, *it will overwrite it*.

--

You can provide a vector as an argument for many functions.
--

```{r Vectors 2}
sqrt(new.object)
```

---

# Character Vectors

We often work with data that are categorical. To create a vector of text elements—**strings** in programming terms—we must place the text in quotes:

```{r}
string.vector <- c("Atlantic", "Pacific", "Arctic")
string.vector
```

--

Categorical data can also be stored as a **factor**, which has an underlying numeric representation. Models will convert factors to dummies.<sup>1</sup>

```{r}
factor.vector <- factor(string.vector)
factor.vector
```

.footnote[[1] Factors have **levels** which you can use to set a reference category in models using `relevel()`.]

---
# Saving and Loading Objects

You can save an R object on your computer as a file to open later:
```{r}
save(new.object, file="new_object.RData")
```

--

You can open saved files in R as well:
```{r}
load("new_object.RData")
```

--
If you are only working within R, `.RData` is a great way of working with data, highly recommend. It uses compression (once reduced a 16 GB file to 2 GB) and removes the need to use packages to import data. 
---

# More Complex Objects

The same principles shown with vectors can be used with more complex objects like **matrices**, **arrays**, **lists**, and **dataframes** (lists which look like matrices but can hold multiple data types at once).

Most data sets you will work with will be read into R and stored as a **dataframe**, so the remainder of this workshop will mainly focus on using these objects.

---
class: inverse, center, middle


# Loading Dataframes

---

# Delimited Text Files

The easiest way to work with external data—that isn't in R format—is for it to be stored in a *delimited* text file, e.g. comma-separated values (**.csv**) or tab-separated values (**.tsv**). 

--

R has a variety of built-in functions for importing data stored in text files, like `read.table()` and `read.csv()`.


By default, these functions will read *character* (string) columns in as a *factor*.

To disable this, use the argument `stringsAsFactors = FALSE`, like so:

```{r, eval=FALSE}
new_df <- read.csv("some_spreadsheet.csv", stringsAsFactors = FALSE)
```

---
# Data from Other Software

Working with **Stata**, **SPSS**, or **SAS** users? You can use a **package** to bring in their saved data files:
 
* `foreign`
    + Part of base R
    + Functions: `read.spss()`, `read.dta()`, `read.xport()`
    + Less complex but sometimes loses some metadata
* `haven`
    + Part of the `tidyverse` family
    + Functions: `read_spss()`, `read_dta()`, `read_sas()`
    + Keeps metadata like variable labels


In general, use `haven`. For less common formats, Google it. I've yet to encounter a data format without an 
R package to handle it (or at least a clever hack).

If you encounter an ambiguous file extension (e.g. `.dat`), try opening it with
a good text editor first (e.g. Atom, Sublime); there's a good chance it is actually raw text
with a delimiter or fixed format that R can handle!

---

# Installing Packages

But what are packages? 

--

Packages contain functions (and sometimes data) created by the community. The real power of R is found in add-on packages!

--

For the remainder of today's course, we will work with data from the `gapminder` package.

These data are a panel data describing 142 countries observed every 5 years from 1952 to 2007.

--

We can install `gapminder` from the Comprehensive R Archive Network (CRAN):

```{r, eval=FALSE}
install.packages("gapminder")
```

You only need to install a package **once** for any given version of R. You need to reinstall packages after upgrading R.

---

# Loading Packages

To load a package, use `library()`:

```{r}
library(gapminder)
```

Once a package is loaded, you can call on functions or data inside it.

```{r}
data(gapminder) # Places data in your global environment
head(gapminder) # Displays first six elements of an object
```
