---
title: "Indexing and basic plots"
author: "Felix Wüthrich"
date: "14.09.2021"
output:
  xaringan::moon_reader:
    css: xaringan-themer.css
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      always_allow_html: true

---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(fig.retina = 3, message = FALSE) #rendering sharp plots / not showing messages
```

```{r xaringan-themer, include=FALSE, warning=FALSE}
library(xaringanthemer)
style_mono_light(base_color = "#081c15")
#class: inverse, center, middle
library(tidyverse)
library(gapminder)
```

# Indices and Dimensions

Dataframes have *2* dimensions: **rows** and **columns**. Square brackets allow us to numerically **subset** in the format of `object[row, column]`. Leaving the row or column place empty selects *all* elements of that dimension.


```{r}
gapminder[1,] # First row
```
```{r}
gapminder[,1] # First column
```

---

```{r}
gapminder[c(1,2,3,4,5), c(3,4)] # First five rows, third and fourth column
```

```{r}
gapminder[1:5, 3:4] # First five rows, third and fourth column
```


---

# Dataframes and Names

Columns in dataframes can also be accessed using their names with the `$` extract operator. This will return the column as a vector:

```{r}
gapminder$gdpPercap[1:10]
```

--

Note here I *also* used brackets to select just the first 10 elements of that column.

You can mix subsetting formats! In this case I provided only a single value (no column index) because **vectors** have *only one dimension* (length).

If you try to subset something and get a warning about "incorrect number of dimensions", check your subsetting!

---

# Indexing by Expression

We can also index using expressions—logical *tests*.

```{r}
gapminder[gapminder$year==1952, ]
```

---

# How Expressions Work

What does `gapminder$year==1952` actually do? 

--

```{r}
head(gapminder$year==1952, 50) # display first 50 elements
```

--

It returns a vector of `TRUE` or `FALSE` values.

When used with the subset operator (`[]`), elements for which a `TRUE` is given are returned while those corresponding to `FALSE` are dropped.

---

# Logical Operators

We used `==` for testing "equals": `gapminder$year==1952`.

--

There are many other [logical operators](http://www.statmethods.net/management/operators.html):

--

* `!=`: not equal to
--

* `>`, `>=`, `<`, `<=`: less than, less than or equal to, etc.
--

* `%in%`: used with checking equal to one of several values

--

Or we can combine multiple logical conditions:

* `&`: both conditions need to hold (AND)
--

* `|`: at least one condition needs to hold (OR)
--

* `!`: inverts a logical condition (`TRUE` becomes `FALSE`, `FALSE` becomes `TRUE`)

--

Logical operators are one of the foundations of programming. You should experiment with these to become familiar with how they work!

---

# Sidenote: Missing Values

Missing values are coded as `NA` entries without quotes:

```{r}
vector_w_missing <- c(1, 2, NA, 4, 5, 6, NA)
```

Even one `NA` "poisons the well": You'll get `NA` out of your calculations unless you remove them manually or use the extra argument `na.rm = TRUE` in some functions:

```{r}
mean(vector_w_missing)
mean(vector_w_missing, na.rm=TRUE)
```

---
# Finding Missing Values

**WARNING:** You can't test for missing values by seeing if they "equal" (`==`) `NA`:

```{r}
vector_w_missing == NA
```

But you can use the `is.na()` function:

```{r}
is.na(vector_w_missing)
```

We can use subsetting to get the equivalent of `na.rm=TRUE`:

```{r}
mean(vector_w_missing[!is.na(vector_w_missing)]) #<<
```

.pull-right[
.footnote[
`!` *reverses* a logical condition. Read the above as "subset *not* `NA`"
]
]

---

# Multiple Conditions Example

Let's say we want observations from Oman after 1980 and through 2000.

--

```{r}
gapminder[gapminder$country == "Oman" &
          gapminder$year > 1980 &
          gapminder$year <= 2000, ]
```

.footnote[
Note we always need to use the full object name in each subseting argument, rather than just `country == "Oman"` alone. You can subset one object using another this way (e.g. `gapminder[other_data$some_variable == x, ]`).
]

---

# Saving a Subset

If we think a particular subset will be used repeatedly, we can save it and give it a name like any other object:

```{r}
China <- gapminder[gapminder$country == "China", ]
head(China, 4)
```

---
# Another Operator: `%in%`

A common thing we may want to do is subset rows to things in some *set*.

We can use `%in%` like `==` but it matches *any element* in the vector on its right. 

```{r}
former_yugoslavia <- c("Bosnia and Herzegovina", "Croatia", 
              "Macedonia", "Montenegro", "Serbia", "Slovenia")
yugoslavia <- gapminder[gapminder$country %in% former_yugoslavia, ]
head(yugoslavia)
```

---
## Create New Columns

We can create new columns (variables) in a dataframe using the same subsetting functions.



```{r}
yugoslavia$pop_million <- yugoslavia$pop / 1000000
yugoslavia$life_exp_past_40 <- yugoslavia$lifeExp - 40
head(yugoslavia)
```

.footnote[
Note one of our new variables is not displayed due to limited width.
]

---
# `ifelse()`


A common function used in general in R programming is `ifelse()`. This returns a value depending on logical tests.

```{r, eval=FALSE}
ifelse(test = x==y, yes = 1, no = 2)
```

Output from `ifelse()`:
* `ifelse()` returns the value assigned to `yes` (in this case, `1`) if `x==y` is `TRUE`.

* `ifelse()` returns `no` (in this case, `2`) if `x==y` is `FALSE`.

* `ifelse()` returns `NA` if `x==y` is neither `TRUE` nor `FALSE`.

Note we can omit explicitly typing function arguments like `test = ` if we enter them in the order of arguments shown in the function's help page.

---
# `ifelse()` Example

.small[
```{r}
yugoslavia$short_country <-  ifelse(yugoslavia$country == "Bosnia and Herzegovina",
                                    "B and H",
                                    as.character(yugoslavia$country))
yugoslavia[yugoslavia$year==1952, c(1,9)] # Selecting just columns 1 and 9
```
]

Read this as "For each row, if `country` equals `"Bosnia and Herzegovina"`, make `short_country` equal to `"B and H"`, otherwise make it equal to that row's original value of `country` (as character, rather than factor, data)."

This is a simple way to change some values but not others!

---

class: middle, inverse, center

## Basic Plots

---

# Histograms

We can use the `hist()` function to generate a histogram of a vector:

```{r, eval = FALSE}
hist(gapminder$lifeExp,
     xlab = "Life Expectancy (years)", #<<
     main = "Observed Life Expectancies of Countries") #<<
```

- `xlab =` is used to set the label of the x-axis of a plot.
- `main = ` is used to set the title of a plot.
- Use `?hist` to see additional options available for customizing a histogram.

---

class: center

```{r, echo=FALSE}
hist(gapminder$lifeExp,
     xlab = "Life Expectancy (years)", #<<
     main = "Observed Life Expectancies of Countries") #<<
```


---

# Scatter Plots

```{r, eval=FALSE}
plot(lifeExp ~ gdpPercap, data = gapminder, #<<
     xlab = "ln(GDP per Capita)",
     ylab = "Life Expectancy (years)",
     main = "Life Expectancy and log GDP per Capita",
     pch = 16, log="x") # log="x" sets x axis to log scale!
abline(h = mean(gapminder$lifeExp), col = "firebrick") #<<
abline(v = mean(gapminder$gdpPercap), col = "cornflowerblue") #<<
```

Note that `lifeExp ~ gdpPercap` is a **formula** of the type `y ~ x`. The first element (`lifeExp`) gets plotted on the y-axis and the second (`gdpPercap`) goes on the x-axis.

The `abline()` calls place horizontal (`h =`) or vertical (`v =`) lines at the means of the variables used in the plot. 

---

class: center

```{r, echo=FALSE}
plot(lifeExp ~ gdpPercap, data = gapminder, #<<
     xlab = "ln(GDP per Capita)",
     ylab = "Life Expectancy (years)",
     main = "Life Expectancy and log GDP per Capita",
     pch = 16, log="x") # log="x" sets x axis to log scale!
abline(h = mean(gapminder$lifeExp), col = "firebrick") #<<
abline(v = mean(gapminder$gdpPercap), col = "cornflowerblue") #<<
```
